\section*{Declaration} Hereby I~declare that this paper is my original authorial work, which
I~have worked out on my own. All sources, references, and literature
used or excerpted during elaboration of this work are properly cited
and listed in complete reference to the due source.

\vspace{1cm}
\begin{flushright}
Petr Kubica
\end{flushright}

\vfill\noindent
\textbf{Advisor:} RNDr. Petr Ročkai, Ph.D.   % XXX
% \\\textbf{Konzultant:} nope % XXX
\cleardoublepage

\section*{Acknowledgements} % XXX
I would like to thank everyone who has helped me make this thesis possible, especially my advisor Petr Ročkai for his guidance and advice. I would also like to thank my friends and colleagues, particularly from the ParaDiSe laboratory, for their support. To anyone negatively affected by my lack of time for other work and activities, I apologize and thank you for your understanding. Finally, I would like to thank my family for their support.
\cleardoublepage

\section*{Abstract} % XXX
This thesis describes the design and implementation of an ahead-of-time compiler for a subset of TypeScript. The compiler is implemented as an extension of Jaculus-machine -- a small, embeddable JavaScript runtime environment. The compiler makes use of TypeScript's type annotations to generate more monomorphic code than what is possible by using JavaScript. It can compile parts of an input program to machine code, which can be used to improve its performance. The compiler is then evaluated on several benchmarks and compared to only using an interpreter.

\subsection*{Keywords} % XXX
JavaScript, TypeScript, compiler, ahead-of-time compilation, embedded systems
\cleardoublepage
