\chapter{Conclusion}


The goal of this thesis was to implement an ahead-of-time compiler for a subset of TypeScript and integrate it into Jaculus-machine.

The presented compiler is able to compile programs written in a subset of TypeScript described in Chapter \ref{chap:supported_lang} to machine code. The supported language is large enough to be usable for writing simple programs and library subroutines.

The compiler also provides a significant performance improvement over interpreted code.

Source code of the compiler is available in the attached archive described in Appendix \ref{app:attachments}.


\section{Limitations and future work}

At present, the compiler supports several 64-bit architectures. This allows it to be used on many devices, including single-board computers like the Raspberry Pi Zero 2. However, most microcontrollers are based on 32-bit architectures or smaller, which are not supported by the compiler.

Because the compiler lacks support for the Xtensa and riscv32 architectures, it cannot be integrated into the Jaculus port for ESP32 microcontrollers -- the primary platform targeted by Jaculus. Extending the MIR backend to support these architectures should be possible with reasonable effort, and will be the next step in the development of the compiler.

The limited language support is another limitation of the compiler, as writing more complex programs without all features of the language is inconvenient. Future work should focus on extending the language support of the compiler or allowing the user to avoid these limitations in some way (e.g., by not invoking the compiler for some files).

We would also like to improve the management of reference-counted values in the compiler. In the current implementation with a cleanup pool, it is possible to write loops, that generate large numbers of values that are saved to the cleanup pool. This can lead to an excessive memory usage. Instead, we would like to implement a more sophisticated mechanism that would decrease reference counts immediately when a value is no longer needed and would not require a cleanup pool.
